Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-11-15 manuscript/manuscript.tex
Fix editor's remark 21
Fix editor's remark 20
Specify smooth function in abstract

\begin{summary}
We present a new methodology to compute the gravitational fields generated by
tesseroids (spherical prisms) whose density varies continuously with depth according to
an arbitrary function.
It approximates the gravitational fields through the Gauss-Legendre Quadrature along
with two discretization algorithms that automatically control its accuracy by adaptively
dividing the tesseroid into smaller ones.
The first one is a preexisting adaptive discretization algorithm that reduces
the errors due to the distance between the tesseroid and the computation
point.
The second is a new density-based discretization algorithm that
decreases the errors introduced by the variation of the density function with depth.
The amount of divisions made by each algorithm is indirectly controlled
by two parameters: the distance-size ratio and the delta ratio.
We have obtained analytical solutions for a spherical shell with radially variable
density and compared them to the results of the numerical model for linear and
exponential density functions.
These comparisons allowed us to obtain optimum values for the distance-size and
delta ratios that yield an accuracy of 0.1\% of the analytical solutions.
The resulting optimal values of distance-size ratio for the gravitational potential, its
gradient, and Marussi tensor are 1, 2 and 8, respectively.
A delta ratio of 0.2 is needed for the computation of the gravitational potential and
its gradient components, while a value of 0.01 must be used for the Marussi tensor
components.
Lastly, we apply this new methodology to model the Neuqu\'en Basin, a foreland basin in
Argentina with a maximum depth of over 5000 m, using an exponential density function.
\end{summary}

@@

The main challenge of the GLQ integration is loss of accuracy when the computation point
gets closer to the tesseroid \citep{Ku1977}.
\citet{Uieda2016} built on the adaptive discretization algorithm of \citet{Li2011}
to automatically obtain integration results with 0.1\% accuracy.
The algorithm consist in recursively splitting the tesseroid into smaller ones when a
threshold is exceeded,
namely when the normalized distance to the computation point is greater than a
``distance-size ratio'' ($D$).
\citet{Uieda2016} have also obtained standard values of $D$
for the gravitational potential, acceleration, and gradient tensor components
by comparing the numerical model with the analytical solution of the fields generated by
a spherical shell.

@@

Consider an arbitrary tesseroid in a geocentric spherical coordinate system defined by
pairs of geocentric latitudinal ($\phi_1$, $\phi_2$), longitudinal ($\lambda_1$,
$\lambda_2$) and radial ($r_1$, $r_2$) boundaries.
We define an external computation point $P(r, \phi, \lambda)$ located at radius $r$,
geocentric latitude $\phi$, and longitude $\lambda$ where the gravitational fields the
tesseroid generates are going to be calculated.
The first- and second-order derivatives of the gravitational potential are taken with
respect to the local north-oriented Cartesian coordinate system of $P$
(Fig.~\ref{fig:tesseroid}).
\citet{Grombein2013} provide efficient formulations for the volume integrals of the
gravitational potential and its first- and second-order derivatives of a tesseroid with
homogeneous density.
Here, we will assume that the tesseroid has a density varying with $r$ according to an
arbitrary function $\rho(r)$.
Thus, the integrals for the gravitational fields are slightly modified to

@@

\iftwocol{
\begin{equation}
    \begin{split}
        \int\limits_{\lambda_1}^{\lambda_2}
        \int\limits_{\phi_1}^{\phi_2}
        \int\limits_{r_1}^{r_2}
        \rho(r') f(r', \phi', \lambda')
        dr' d\phi' d\lambda' \approx& \\
        A
        \sum\limits_{i=1}^{N^r}
        \sum\limits_{j=1}^{N^\phi}
        \sum\limits_{k=1}^{N^\lambda}
        W_i^r W_j^\phi W_k^\lambda
        &\rho(r_i) f(r_i, \phi_j, \lambda_k),
    \end{split}
\label{eq:glq-var-dens}
\end{equation}
}{
\begin{equation}
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') f(r', \phi', \lambda')
    dr' d\phi' d\lambda' \approx
    A
    \sum\limits_{i=1}^{N^r}
    \sum\limits_{j=1}^{N^\phi}
    \sum\limits_{k=1}^{N^\lambda}
    W_i^r W_j^\phi W_k^\lambda \rho(r_i) f(r_i, \phi_j, \lambda_k),
\label{eq:glq-var-dens}
\end{equation}
}

@@

\noindent $f(r', \phi', \lambda')$ is the non-density dependent portion of the
integration kernels, $(r_i, \phi_j, \lambda_k)$ are the quadrature nodes, and
$W_i^r$, $W_j^\phi$, $W_k^\lambda$ are the quadrature weights.

>>>>

\begin{summary}
We present a new methodology to compute the gravitational fields generated by
tesseroids (spherical prisms) whose density varies continuously with depth according to
an arbitrary smooth function.
It approximates the gravitational fields through the Gauss-Legendre Quadrature along
with two discretization algorithms that automatically control its accuracy by adaptively
dividing the tesseroid into smaller ones.
The first one is a preexisting adaptive discretization algorithm that reduces
the errors due to the distance between the tesseroid and the computation
point.
The second is a new density-based discretization algorithm that
decreases the errors introduced by the variation of the density function with depth.
The amount of divisions made by each algorithm is indirectly controlled
by two parameters: the distance-size ratio and the delta ratio.
We have obtained analytical solutions for a spherical shell with radially variable
density and compared them to the results of the numerical model for linear and
exponential density functions.
These comparisons allowed us to obtain optimum values for the distance-size and
delta ratios that yield an accuracy of 0.1\% of the analytical solutions.
The resulting optimal values of distance-size ratio for the gravitational potential, its
gradient, and Marussi tensor are 1, 2 and 8, respectively.
A delta ratio of 0.2 is needed for the computation of the gravitational potential and
its gradient components, while a value of 0.01 must be used for the Marussi tensor
components.
Lastly, we apply this new methodology to model the Neuqu\'en Basin, a foreland basin in
Argentina with a maximum depth of over 5000 m, using an exponential density function.
\end{summary}

@@

The main challenge of the GLQ integration is loss of accuracy when the computation point
gets closer to the tesseroid \citep{Ku1977}.
\citet{Uieda2016} built on the adaptive discretization algorithm of \citet{Li2011}
to automatically obtain integration results with 0.1\% accuracy.
The algorithm consist in recursively splitting the tesseroid into smaller ones when a
threshold is exceeded,
namely when the normalized distance to the computation point is greater than a
``distance-size ratio'' ($D$).
\citet{Uieda2016} have also obtained standard values of $D$
for the gravitational potential, acceleration, and gradient tensor components
by comparing the numerical model with the fields generated by a spherical shell.

@@

We define an external computation point $P(r, \phi, \lambda)$ in a geocentric spherical
coordinate system at a radius $r$, geocentric latitude $\phi$, and longitude $\lambda$
where the gravitational fields are going to be calculated.
The first and second derivatives of the gravitational potential are taken with
respect to the local north-oriented Cartesian coordinate system of $P$
(Fig.~\ref{fig:tesseroid}).
\citet{Grombein2013} provide efficient formulations for the volume integrals of the
gravitational potential and its first and second derivatives of a tesseroid with
homogeneous density.
Here, we will assume that the tesseroid has a density varying with $r$ according to an
arbitrary function $\rho(r)$.
Thus, the integrals for the gravitational fields are slightly modified to

@@

\iftwocol{
\begin{equation}
    \begin{split}
        \iiint\limits_\Omega \rho(r') f(r', \phi', \lambda')
        d\Omega \approx& \\
        A
        \sum\limits_{i=1}^{N^r}
        \sum\limits_{j=1}^{N^\phi}
        \sum\limits_{k=1}^{N^\lambda}
        & W_i^r W_j^\phi W_k^\lambda \rho(r_i) f(r_i, \phi_j, \lambda_k),
    \end{split}
\label{eq:glq-var-dens}
\end{equation}
}{
\begin{equation}
    \iiint\limits_\Omega \rho(r') f(r', \phi', \lambda') d\Omega \approx
    A
    \sum\limits_{i=1}^{N^r}
    \sum\limits_{j=1}^{N^\phi}
    \sum\limits_{k=1}^{N^\lambda}
    W_i^r W_j^\phi W_k^\lambda \rho(r_i) f(r_i, \phi_j, \lambda_k),
\label{eq:glq-var-dens}
\end{equation}
}

@@

\noindent $(r_i, \phi_j, \lambda_k)$ are the quadrature nodes, and $W_i^r$ are the
quadrature weights.
